% =====================================================================
%  Network Intrusion Detection on CICIDS2017
%  Springer LNCS / CCIS format  (target venues: ICTA 2026, FDSE 2026)
%  Compiled FREE on GitHub Actions via xu-cheng/latex-action@v3
%
%  GOLDEN RULE: every number in this paper must come from the Colab
%  experiments (experiments/ids_cicids2017.ipynb). Placeholders are
%  marked TODO and must be replaced with real measured values.
% =====================================================================
\documentclass[runningheads]{llncs}
% Round 1 (double-blind): keep \anontrue -> hides authors.
% Round 2 (camera-ready): change to \anonfalse -> shows full author info.
\newif\ifanon\anontrue

\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{amsmath}
\usepackage{multirow}
\usepackage{array}
\usepackage{url}
\usepackage{tikz}
\usetikzlibrary{positioning,fit,arrows.meta}
% NOTE: hyperref intentionally NOT loaded — it clashes with llncs
% (\@xdblarg error). Springer adds hyperlinking for the final version.

\begin{document}

\title{Network Intrusion Detection on CICIDS2017: A Comparative Study of
Machine Learning and Deep Learning Models with a Cloud-Native
Deployment Design}

\ifanon
\author{Anonymous Author(s)}
\institute{Affiliation withheld for double-blind review}
\else
\author{Huy-Thanh Pham\inst{1} \and Tuan-Anh Truong\inst{1} \and
Vinh-Khuong Truong\inst{2}}
\authorrunning{H.-T. Pham et al.}
\institute{Faculty of Computer Science and Engineering,\\
Ho Chi Minh City University of Technology (HCMUT), VNU-HCM, Vietnam \\
\email{huythanh@clouds.io.vn, anhtt@hcmut.edu.vn}
\and Ton Duc Thang University, Ho Chi Minh City, Vietnam \\
\email{trvinhkhuong@gmail.com}}
\fi

\maketitle

\begin{abstract}
Network intrusion detection is a cornerstone of modern cyber-defense, yet
the transition from an accurate offline classifier to a scalable,
continuously serving system is frequently overlooked. In this paper we
present a reproducible comparative study of classical machine learning and
deep learning models for network intrusion detection on the widely used
CICIDS2017 benchmark, with cross-dataset validation on UNSW-NB15. We design
a unified pipeline covering data cleaning,
feature selection, and class-imbalance handling, and we evaluate Logistic
Regression, Random Forest, XGBoost, and a multilayer perceptron under a
single protocol using accuracy, precision, recall, macro-F1, and ROC-AUC.
Beyond predictive quality, we quantify training and inference cost so that
model selection reflects deployability rather than accuracy alone. Finally,
we propose a cloud-native deployment design that packages the selected
model as a containerized microservice with an automated CI/CD pipeline.
The best model (XGBoost) attains a macro-F1 of 99.2\% with an inference cost
of only 0.006~ms per sample, indicating both high detection quality and low
serving cost.
Our study offers a practical, cost-aware reference for building deployable
intrusion detection systems.
\keywords{Intrusion Detection \and CICIDS2017 \and Machine Learning \and
Deep Learning \and Cloud-Native \and Network Security.}
\end{abstract}

% =====================================================================
\section{Introduction}
\label{sec:intro}
Enterprise and telecommunication networks are exposed to an ever-growing
volume and diversity of attacks, ranging from volumetric denial-of-service
campaigns to stealthy infiltration attempts. The scale and velocity of
modern traffic make purely manual analysis infeasible and motivate
automated, learning-based intrusion detection systems (IDS) that can flag
malicious flows in near real time~\cite{buczak2016survey}. Over the past
decade, machine learning (ML) and, more recently, deep learning (DL)
methods have become the dominant approach to this problem.

Despite substantial progress, two practical gaps persist. First, many
studies report high accuracy on a single model and dataset configuration,
making fair comparison across model families difficult because
preprocessing, feature selection, and evaluation protocols differ from one
work to another. Second, the literature emphasizes offline predictive
quality while paying limited attention to \emph{deployability}: the
computational cost of training and inference, and the engineering path from
a trained artifact to a scalable, continuously operated service. In an
operational setting, a marginally more accurate model that is an order of
magnitude more expensive to serve may be the inferior choice.

\paragraph{Contributions.} This paper addresses these gaps with the
following contributions:
\begin{itemize}
  \item A reproducible preprocessing and evaluation pipeline for the
        CICIDS2017 and UNSW-NB15 datasets, released openly to support
        replication.
  \item A unified comparison of classical ML (Logistic Regression, Random
        Forest, XGBoost) and a deep learning model, evaluated jointly on
        accuracy, precision, recall, macro-F1, and ROC-AUC, together with
        training and inference cost.
  \item A cloud-native deployment design that serves the selected model as
        a containerized microservice driven by an automated CI/CD pipeline.
\end{itemize}

\paragraph{Organization.}
Section~\ref{sec:related} reviews related work.
Section~\ref{sec:data} describes the dataset and preprocessing.
Section~\ref{sec:method} details the models and the evaluation protocol.
Section~\ref{sec:results} reports and discusses the results.
Section~\ref{sec:deploy} presents the deployment design, and
Section~\ref{sec:conclusion} concludes and outlines future work.

% =====================================================================
\section{Related Work}
\label{sec:related}
Machine learning and data mining techniques have been applied extensively
to intrusion detection, spanning supervised classifiers, unsupervised
anomaly detectors, and hybrid schemes; the survey of Buczak and
Guven~\cite{buczak2016survey} provides a broad overview of methods and the
trade-offs between detection accuracy, interpretability, and computational
cost.

Benchmark datasets have strongly shaped progress in the field. The earlier
KDD'99 corpus and its refined variant NSL-KDD~\cite{tavallaee2009nslkdd}
exposed and partially corrected problems such as redundant records and
unrealistic class balance, but both remain dated with respect to modern
traffic and attack types. UNSW-NB15~\cite{moustafa2015unswnb15} introduced
a more contemporary feature set and attack taxonomy. CICIDS2017
~\cite{sharafaldin2018cicids} further improved realism by capturing several
days of labeled traffic that mixes benign behavior with a range of current
attack families, which is why we adopt it as our primary benchmark.

While many works report strong results on these datasets, comparatively few
combine (i) a fair, single-protocol comparison across model families with
(ii) an explicit treatment of computational cost and a concrete deployment
path. Recent studies confirm the continued relevance of CICIDS2017 and of
strong tabular models: Talukder et al.~\cite{talukder2024ml} combine
oversampling with stacking-based feature embedding to handle large and
imbalanced traffic; Almuhanna and Dardouri~\cite{almuhanna2025dl} contrast
deep learning and classical models for anomaly-based detection; and Xu and
Liu~\cite{xu2025robust} systematically evaluate machine learning models on
CICIDS2017. Our work complements these by jointly weighing detection quality
against serving cost and by pairing the comparison with a deployable,
cloud-native design.

% =====================================================================
\section{Dataset and Preprocessing}
\label{sec:data}
The CICIDS2017 dataset~\cite{sharafaldin2018cicids} contains roughly five
days of labeled network flows, comprising benign traffic together with
several attack families such as denial-of-service, distributed
denial-of-service, brute-force, web attacks, and infiltration. Each flow is
described by more than eighty statistical features derived from packet and
flow-level measurements. The data are distributed as multiple CSV files,
which we merge into a single corpus; for computational tractability we draw
a fixed 20\% stratified random sample (seed~42) from each file.

\subsection{Cleaning and Encoding}
We first normalize column names by stripping leading and trailing
whitespace, then replace infinite values with missing markers and drop
records containing missing values, together with exact duplicates that
would otherwise bias evaluation. We cast the task as binary classification,
encoding every non-benign flow as the positive (attack) class and benign
flows as the negative class. Numeric features are standardized to zero mean
and unit variance using statistics estimated on the training split only, to
avoid information leakage.

\subsection{Feature Selection}
To reduce dimensionality and inference cost, we rank features by importance
using a Random Forest~\cite{breiman2001random} trained on the training
split and retain the top $k=20$ features. This keeps the most informative
attributes while discarding redundant or low-signal columns.

\subsection{Class Imbalance}
Benign flows substantially outnumber attack flows, which can bias a
classifier toward the majority class. We mitigate this on the training
split using the Synthetic Minority Over-sampling Technique
(SMOTE)~\cite{chawla2002smote}; the test split is left untouched so that
reported metrics reflect the original distribution. After cleaning, the
working set contains 530{,}222 flows with 78 numeric features and an attack
ratio of 17.85\%, split 70/30 into stratified training and test partitions.

% =====================================================================
\section{Methodology}
\label{sec:method}
\subsection{Models}
We compare four models that span a representative range of capacity and
cost. \emph{Logistic Regression} serves as a linear baseline. \emph{Random
Forest}~\cite{breiman2001random} is a strong bagging ensemble of decision
trees. \emph{XGBoost}~\cite{chen2016xgboost} is a gradient-boosted tree
method that is widely competitive on tabular data. Finally, a
\emph{multilayer perceptron} (MLP) represents the deep learning family with
two hidden layers.

\subsection{Experimental Setup}
All experiments are conducted on the free tier of Google Colab using
scikit-learn and XGBoost. We use a single stratified 70/30 train/test
split. A fixed random seed (42) is used throughout to ensure
reproducibility, and the complete pipeline is provided in the accompanying
notebook.

\subsection{Evaluation Metrics}
Let $TP$, $TN$, $FP$, and $FN$ denote true positives, true negatives,
false positives, and false negatives. We report
\begin{equation}
\mathrm{Precision}=\frac{TP}{TP+FP},\qquad
\mathrm{Recall}=\frac{TP}{TP+FN},
\end{equation}
\begin{equation}
\mathrm{F1}=2\cdot\frac{\mathrm{Precision}\cdot\mathrm{Recall}}
{\mathrm{Precision}+\mathrm{Recall}},
\end{equation}
together with overall accuracy and the macro-averaged F1, which weights the
two classes equally and is therefore informative under class imbalance. We
additionally report the area under the ROC curve (ROC-AUC). To capture
deployability, we measure wall-clock training time and average inference
time per sample for each model.

% =====================================================================
\section{Experiments and Results}
\label{sec:results}
Table~\ref{tab:results} reports the detection performance of the four
models on the held-out test set, and Table~\ref{tab:cost} reports their
computational cost.

\begin{table}[t]
\centering
\caption{Detection performance on CICIDS2017 (test set). Best value per
column in bold.}
\label{tab:results}
\begin{tabular}{lccccc}
\toprule
Model & Accuracy & Precision & Recall & macro-F1 & ROC-AUC \\
\midrule
Logistic Regression & 0.9186 & 0.7324 & 0.8571 & 0.8697 & 0.9629 \\
Random Forest       & 0.9935 & 0.9719 & \textbf{0.9924} & 0.9890 & 0.9983 \\
XGBoost             & \textbf{0.9952} & \textbf{0.9811} & 0.9921 & \textbf{0.9918} & \textbf{0.9998} \\
Deep model (MLP)    & 0.9763 & 0.8911 & 0.9882 & 0.9613 & 0.9989 \\
\bottomrule
\end{tabular}
\end{table}

\begin{table}[t]
\centering
\caption{Computational cost (training time and per-sample inference).}
\label{tab:cost}
\begin{tabular}{lcc}
\toprule
Model & Train time (s) & Inference (ms/sample) \\
\midrule
Logistic Regression & 10.55  & 0.0001 \\
Random Forest       & 185.76 & 0.0090 \\
XGBoost             & 18.58  & 0.0057 \\
Deep model (MLP)    & 675.54 & 0.0232 \\
\bottomrule
\end{tabular}
\end{table}

Across the four models, the best macro-F1 is achieved by XGBoost at
99.18\,\%, closely followed by Random Forest (98.90\,\%); the linear
baseline (Logistic Regression, 86.97\,\%) trails the tree-based ensembles,
as expected on this largely non-linear feature space. When cost is taken
into account (Table~\ref{tab:cost}), XGBoost offers the most favorable
accuracy-to-cost trade-off: it matches or exceeds Random Forest in accuracy
while training in 18.58\,s versus 185.76\,s, and its inference cost is only
0.0057\,ms per sample versus 0.0232\,ms for the deep model. This supports
our central argument that model selection for an operational IDS should
weigh serving cost alongside accuracy.

To verify that these results are not an artifact of the particular
train/test partition, we additionally perform 5-fold stratified
cross-validation for the best model on the training data: XGBoost attains a
macro-F1 of $0.9918 \pm 0.0004$, confirming the stability of the reported
performance. Figure~\ref{fig:roc} shows its ROC curve and
Table~\ref{tab:cm} the confusion matrix on the test set, where only 542
false positives and 224 false negatives occur out of roughly 159k flows.

\begin{table}[t]
\centering
\caption{Confusion matrix of the XGBoost detector on the CICIDS2017 test set.}
\label{tab:cm}
\begin{tabular}{ll cc}
\toprule
 & & \multicolumn{2}{c}{Predicted} \\
\cmidrule(lr){3-4}
 & & Benign & Attack \\
\midrule
\multirow{2}{*}{Actual} & Benign & 130129 & 542 \\
 & Attack & 224 & 28172 \\
\bottomrule
\end{tabular}
\end{table}

\begin{figure}[t]
\centering
\begin{tikzpicture}[scale=4.2]
\draw[->] (0,0) -- (1.08,0) node[right,font=\small] {FPR};
\draw[->] (0,0) -- (0,1.08) node[above,font=\small] {TPR};
\foreach \x in {0,0.5,1} \draw (\x,0) -- (\x,-0.02) node[below,font=\scriptsize] {\x};
\foreach \y in {0.5,1} \draw (0,\y) -- (-0.02,\y) node[left,font=\scriptsize] {\y};
\draw[gray,dashed] (0,0) -- (1,1);
\draw[blue,very thick]
  (0,0) -- (0,0.8023) -- (0.0317,0.9997) -- (0.0709,0.9999) --
  (0.0973,0.9999) -- (0.1369,1.0) -- (1.0,1.0);
\end{tikzpicture}
\caption{ROC curve of the XGBoost detector on CICIDS2017 (AUC $=0.9998$).}
\label{fig:roc}
\end{figure}

The relatively lower precision of the linear baseline (0.7324) reflects a
higher false-positive rate on minority attack classes that share
statistical characteristics with benign traffic; the tree-based ensembles
largely mitigate this, with XGBoost reaching a precision of 0.9811.
Reducing the residual false negatives on under-represented attack families
remains the principal avenue for improvement.

\paragraph{Cross-dataset validation.}
To assess whether our findings generalize beyond a single benchmark, we
repeat the comparison on the UNSW-NB15 dataset~\cite{moustafa2015unswnb15}
(175{,}341 flows, 42 features, attack ratio 68\%) under the same pipeline
and a stratified 70/30 split. Table~\ref{tab:unsw} reports the results.
Although the absolute scores are lower on this more balanced and harder
dataset, the model ranking is preserved: XGBoost again achieves the best
macro-F1 (0.9497) and ROC-AUC (0.9930), closely followed by Random Forest,
with the linear baseline trailing. The consistency of this ordering across
two independent datasets indicates that our cost-aware conclusion is not an
artifact of a single benchmark.

\begin{table}[t]
\centering
\caption{Detection performance on the UNSW-NB15 dataset (test set). Best
value per column in bold.}
\label{tab:unsw}
\begin{tabular}{lccccc}
\toprule
Model & Accuracy & Precision & Recall & macro-F1 & ROC-AUC \\
\midrule
Logistic Regression & 0.8879 & 0.9108 & 0.9260 & 0.8698 & 0.9331 \\
Random Forest       & 0.9554 & 0.9678 & 0.9666 & 0.9487 & 0.9925 \\
XGBoost             & \textbf{0.9563} & 0.9667 & \textbf{0.9692} & \textbf{0.9497} & \textbf{0.9930} \\
Deep model (MLP)    & 0.9361 & \textbf{0.9757} & 0.9293 & 0.9284 & 0.9893 \\
\bottomrule
\end{tabular}
\end{table}

% =====================================================================
\section{Cloud-Native Deployment Design}
\label{sec:deploy}
To close the gap between an offline classifier and an operational detector,
we package the selected model as a stateless REST microservice. The trained
artifact is exported and wrapped in a lightweight inference service
(implemented with FastAPI), containerized with Docker, and deployed on
Kubernetes, where horizontal replicas provide scalability and resilience.
A CI/CD pipeline automates the build, test, and deployment stages so that a
new model version can be validated and rolled out without manual
intervention; this mirrors the continuous-delivery practice expected of
cloud-native systems and is reproduced for the paper artifact itself, whose
PDF is compiled by a hosted CI runner.

Figure~\ref{fig:arch} summarizes the serving architecture. The selected
XGBoost model has a measured inference cost of 0.0057~ms per sample,
corresponding to a single-thread throughput on the order of $10^5$
predictions per second; this indicates that the detector comfortably meets
near-real-time requirements, while a full end-to-end load test of the
deployed service under concurrent traffic is left for future work.

\begin{figure}[t]
\centering
\begin{tikzpicture}[
  >={Stealth[length=2mm]},
  box/.style={draw, rounded corners, align=center, font=\small,
              minimum height=9mm, inner sep=5pt},
  node distance=10mm and 12mm]
  \node[box] (flows) {Network flows\\(CICIDS2017)};
  \node[box, right=of flows] (pre) {Preprocessing\\and features};
  \node[box, right=of pre] (svc) {IDS microservice\\(FastAPI, Docker)};
  \node[box, right=of svc] (out) {Alerts /\\dashboard};
  \node[box, above=10mm of svc] (cicd) {CI/CD pipeline\\(build--test--deploy)};
  \draw[->] (flows) -- (pre);
  \draw[->] (pre) -- (svc);
  \draw[->] (svc) -- (out);
  \draw[->, dashed] (cicd) -- (svc);
  \node[draw, dotted, rounded corners, inner sep=9pt,
        fit=(pre)(svc)(out),
        label={[font=\small\itshape]below:Kubernetes cluster}] {};
\end{tikzpicture}
\caption{Cloud-native serving architecture: the selected model is exposed as a
FastAPI microservice, containerized with Docker, deployed on a Kubernetes
cluster, and released through an automated CI/CD pipeline.}
\label{fig:arch}
\end{figure}

% =====================================================================
\section{Conclusion and Future Work}
\label{sec:conclusion}
We presented a reproducible, cost-aware comparison of machine learning and
deep learning models for network intrusion detection on CICIDS2017 and
UNSW-NB15, paired with a cloud-native deployment design. Our results
(Tables~\ref{tab:results}--\ref{tab:cost}) indicate that XGBoost provides
the best balance of detection quality (99.18\,\% macro-F1) and serving cost
(18.58\,s training, 0.0057\,ms per-sample inference), reinforcing the view
that deployability should be a first-class criterion in IDS model selection.

Several directions extend this work. Online and continual learning would
allow the detector to adapt to concept drift as traffic and attack patterns
evolve. Explainable detection would help analysts triage alerts and build
trust in automated decisions. Graph-based and sequence models could exploit
relational and temporal structure beyond per-flow features, and a federated
formulation would enable collaborative detection across distributed
telecommunication nodes without centralizing sensitive data. These
constitute the intended trajectory of our continuing doctoral research.

\ifanon\else
\subsubsection*{Acknowledgements.} % TODO (tuỳ chọn): lời cảm ơn / nguồn tài trợ.
\fi

\bibliographystyle{splncs04}
\bibliography{references}

\end{document}
